---
title: "Data Manipulation"
author: "Mark Dunning"
date: "20/11/2014"
output: beamer_presentation
---

##Calculating new variables

```{r}
data <- read.csv("data/ozone.csv")
TempCelc <- (data$Temp - 32)/1.8
plot(TempCelc)
```

## Revisit the ozone dataset

```{r}
data <- read.csv("data/ozone.csv")
```

##Re-ordering and sorting

- At the moment, these data are in date-order

```{r}
head(data)
```
- We might want to know the hottest days

##Re-ordering and sorting

```{r}
sort(data$Temp)
```

##Re-ordering and sorting

```{r}
sort(data$Temp,decreasing = TRUE)
```

##Re-ordering and sorting

- What is the difference between the output of `sort` and `order`?
```{r}
tempOrder <- order(data$Temp, decreasing = TRUE)
length(tempOrder)
tempOrder
```

##Re-ordering and sorting

- `sort` gives the *values* in sorted order
- `order` gives *indices*
- we can use the result of `order` to subset the data
```{r} 
tempOrder[1:5]
data[tempOrder[1:5],]
```

##Writing a new file

- At this point, we might want to write our re-ordered data to a file

```{r}
newData <- data[tempOrder,]
dim(newData)
write.csv(newData, file="reorderedWeather.csv")
```


##General Subsetting

- We have already seen how to subset using numeric indexes
- We can also subset using *logical* vectors
- i.e. a vector of `TRUE` or `FALSE` values
- The `TRUE` or `FALSE` values can be derived from a test such as
  - `<`, `>`, `==`
- Mulitple conditions can be tested using `&` (`and`)  `|` (*or*)


## Adding points

- Suppose we are interested in days with Ozone level over 100
- We can form a *logical* vector and get the `TRUE` values using the `which` function

```{r}
highOzone <- which(data$Ozone > 100)
data[highOzone,]
```

## Adding points

- The `points` funtion is used to add points to an existing plot
- We need to give it a set of *x* and *y* coordinates
- The *x* values are the indices we've just computed. 
- *y* values are obtained by subsetting the *Ozone* variable

```{r}
newX <- highOzone
newY <- data$Ozone[newX]
```

## Adding points


```{r fig.width=8,fig.height=4}
highOzone <- which(data$Ozone > 100)
plot(data$Ozone)
abline(h=100)
points(newX, newY,col="red",pch=16)
```


##Subsetting by text

We now consider the clinical characteristics of a breast cancer cohort

```{r}
clinical <- read.delim("data/NKI295.pdata.txt")
```

##Subsetting by text

We might wish to know the identity of *ER negative* samples

- Note the double `==`

```{r}

clinical$ER == "Negative"

```

##Returning indices

```{r}
which(clinical$ER == "Negative")
```

##Potential trap number 1.

```{r}
which(clinical$ER == "negative")
```

##Potential trap number 2.


```{r}
which(clinical$er == "Negative")
```

##Potential trap number 3.

```{r}
match("Negative", clinical$ER)
```


##Use in subsetting

```{r}
clinical[which(clinical$ER == "Negative"),]
```

##Alternative

```{r}
clinical[grep("Negative",clinical$ER),]
```

##Match multiple strings

```{r}
clinical[which(clinical$Fan.nearest.centroid %in% 
                 c("Basal", "HER2")),]
```


##Useful functions for manipulating text

- `substr` 
```{r}
substr(clinical$sampleNames,1,3)[1:5]
substr(clinical$sampleNames,1,3)[1:5] == "NKI"
```



##Useful functions for manipulating text

- `strsplit`

```{r}
strsplit(as.character(clinical$sampleNames), "_")[[1]]
matrix(unlist(strsplit(as.character(clinical$sampleNames), "_"))
       ,ncol=2,byrow=TRUE)
```


##Useful functions for manipulating text

Not an extensive list

- `toupper`, `tolower` - convert upper / lower case
- `gsub` -  substitute text
- `paste` - combine text


##Combining data from files

- Each row corresponds to a *gene*
- Each column is a *sample*

```{r}
evalues <- read.delim("data/NKI295.exprs.txt")
dim(evalues)
evalues[1:5,1:5]
```

##Clinical data

- Each row is a *sample*
- Each column is a different *clinical* variable

```{r}
clindata <- read.delim("data/NKI295.pdata.txt")
clindata[1:5,1:5]
```

##Gene annotation

- Each row is a *gene* in the experiment
- Each column is *annotation* about that gene

```{r}
annodata <- read.delim("data/NKI295.fdata.txt")
annodata[1:5,1:5]
```

##Example analyses

- Extract the expression values for a given gene
- Extract the arrays representing a particular clinical subgroup
- Plot gene expression against particular clinical variables
- Compare expression of one gene against another