---
title: "Data Manipulation"
author: "Mark Dunning"
date: "20/11/2014"
output: beamer_presentation
---

##Calculating new variables

```{r}
data <- read.csv("data/ozone.csv")
TempCelc <- (data$Temp - 32)/1.8
plot(TempCelc)
```


##General Subsetting

- We have already seen how to subset using numeric indexes
- We can also subset using *logical* vectors
- i.e. a vector of `TRUE` or `FALSE` values
- The `TRUE` or `FALSE` values can be derived from a test such as
  - `<`, `>`, `==`
- Mulitple conditions can be tested using `&` (`and`)  `|` (*or*)

## Revisit the ozone dataset

```{r}
data <- read.csv("data/ozone.csv")
```

## Adding points

- Suppose we are interested in days with Ozone level over 100
- We can form a *logical* vector and get the `TRUE` values using the `which` function

```{r}
highOzone <- which(data$Ozone > 100)
data[highOzone,]
```

## Adding points

- The `points` funtion is used to add points to an existing plot
- We need to give it a set of *x* and *y* coordinates
- The *x* values are the indices we've jsut computed. 
- *y* values are obtained by subsetting the *Ozone* variable

```{r}
newX <- highOzone
newY <- data$Ozone[newX]
```

## Adding points

- We use a different colour and character for the points. More on this later

```{r fig.width=8,fig.height=4}
highOzone <- which(data$Ozone > 100)
plot(data$Ozone)
abline(h=100)
points(newX, newY,col="red",pch=16)
```

##Combining data from files

- Each row corresponds to a *gene*
- Each column is a *sample*

```{r}
evalues <- read.delim("data/NKI295.exprs.txt")
dim(evalues)
evalues[1:5,1:5]
```

##Clinical data

- Each row is a *sample*
- Each column is a different *clinical* variable

```{r}
clindata <- read.delim("data/NKI295.pdata.txt")
clindata[1:5,1:5]
```

##Gene annotation

- Each row is a *gene* in the experiment
- Each column is *annotation* about that gene

```{r}
annodata <- read.delim("data/NKI295.fdata.txt")
annodata[1:5,1:5]
```

##Example analyses

- Extract the expression values for a given gene
- Extract the arrays representing a particular clinical subgroup
- Plot gene expression against particular clinical variables
- Compare expression of one gene against another