---
title: 'R Graphics:  worked examples'
author: "Mark Dunning"
date: "09/09/2014"
output: pdf_document
toc: yes
---
#Introduction

This document describes how basic statistical analysis and graphical displays can be performed using R. 

The different data types are categorised according to the different options available in GraphPad Prism.

# XY data

##Linear Regression - comparing slopes

###Data exploration

Example data for this example is given in the file `Linear regression.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Linear regression.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

You many notice that a number of entries are listed `NA`. This simply means that these observations are *missing* and should be excluded from numerical calculations. 

The generic `plot` function is able to make a scatter plot of one variable against another. By default we specify the *vector* we want to plot on the x-axis first, followed by a *vector* to be displayed on the y-axis. To extract vectors from our data frame, we can use the `[row,column]` notation, or specify the name of the column with a `$` sign. We will start by the `Control` column against number of minutes. 

```{r}
plot(data$Minutes, data$Control)
```

The default settings will plot open circles at every *x* and *y* position specified by the two input vectors. If we want a line through the coordinates, we can modify the `type` argument to `"l"`.

```{r}
plot(data$Minutes, data$Control,type="l")
```

Or we can have a combination of points and a line by setting `type="b"`

```{r}
plot(data$Minutes, data$Control,type="b")
```

We can put additional lines on the plot, in this case corresponding to other experiments, using the `lines` function. This function assumes that a plot has already been created. We can add a line at a time by specifying different columns from the data frame as our *y* axis in the `lines` function.

```{r}
plot(data$Minutes, data$Control,type="l")
lines(data$Minutes, data$Control.1)
lines(data$Minutes, data$Control.2)
lines(data$Minutes, data$Treated)
lines(data$Minutes, data$Treated.1)
lines(data$Minutes, data$Treated.2)
```

You will notice however that some of the lines go outside the plotting area. This is because the range of y values for the plot is defined using the range of the first line that we plotted (the `data$Control` vector). We can specify a particular range of values using the `ylim` argument to `plot`. A sensible value to use would be the *maximum* value in our data frame. However, there is a problem when we try and calculate this.

```{r}
max(data)
```
Because we have `NA` values in the data frame, the maximum value cannot be determined. That is, unless, we use the argument `na.rm=TRUE`

```{r}
max(data,na.rm=TRUE)
```

We can now specify the y limits for the plot using the `ylim` argument. This argument needs to be a vector (hence we use the `c` function to construct it) consisting the minimum and maximum y values for the plot.

```{r}
plot(data$Minutes, data$Control,type="l",ylim=c(0,max(data,na.rm=TRUE)))
lines(data$Minutes, data$Control.1)
lines(data$Minutes, data$Control.2)
lines(data$Minutes, data$Treated)
lines(data$Minutes, data$Treated.1)
lines(data$Minutes, data$Treated.2)
```

We have managed to display the raw data on a plot, but it is unlikely to win any prizes for being aesthetically pleasing! A couple of obvious modifications would be to add a proper x and y label, and to colour the lines appropriately. To change the labels we can specify the `xlab`, `ylab` and `main` arguments. 

```{r}
plot(data$Minutes, data$Control,type="l",ylim=c(0,max(data,na.rm=TRUE)),col="blue",xlab="Minutes",ylab="Count",main="Variation of counts....")
lines(data$Minutes, data$Control.1,col="blue")
lines(data$Minutes, data$Control.2,col="blue")
lines(data$Minutes, data$Treated,col="red")
lines(data$Minutes, data$Treated.1,col="red")
lines(data$Minutes, data$Treated.2,,col="red")
```

##Nonlinear regression


Example data for this example is given in the file `Exponential decay.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Exponential decay.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```



You many notice that a number of entries are listed `NA`. This simply means that these observations are *missing* and should be excluded from numerical calculations. 

The generic `plot` function is able to make a scatter plot of one variable against another. By default we specify the *vector* we want to plot on the x-axis first, followed by a *vector* to be displayed on the y-axis. To extract vectors from our data frame, we can use the `[row,column]` notation, or specify the name of the column with a `$` sign. We will start by the `Control` column against number of minutes. 

```{r}
plot(data$Minutes, data$Control)
```

The default settings will plot open circles at every *x* and *y* position specified by the two input vectors. If we want a line through the coordinates, we can modify the `type` argument to `"l"`.

```{r}
plot(data$Minutes, data$Control,type="l")
```

Or we can have a combination of points and a line by setting `type="b"`

```{r}
plot(data$Minutes, data$Control,type="b")
```

We can put additional lines on the plot, in this case corresponding to other experiments, using the `lines` function. This function assumes that a plot has already been created. We can add a line at a time by specifying different columns from the data frame as our *y* axis in the `lines` function.

```{r}
plot(data$Minutes, data$Control,type="l")
lines(data$Minutes, data$Control.1)
lines(data$Minutes, data$Control.2)
lines(data$Minutes, data$Treated)
lines(data$Minutes, data$Treated.1)
lines(data$Minutes, data$Treated.2)
```

You will notice however that some of the lines go outside the plotting area. This is because the range of y values for the plot is defined using the range of the first line that we plotted (the `data$Control` vector). We can specify a particular range of values using the `ylim` argument to `plot`. A sensible value to use would be the *maximum* value in our data frame. However, there is a problem when we try and calculate this.

```{r}
max(data)
```
Because we have `NA` values in the data frame, the maximum value cannot be determined. That is, unless, we use the argument `na.rm=TRUE`

```{r}
max(data,na.rm=TRUE)
```

We can now specify the y limits for the plot using the `ylim` argument. This argument needs to be a vector (hence we use the `c` function to construct it) consisting the minimum and maximum y values for the plot.

```{r}
plot(data$Minutes, data$Control,type="l",ylim=c(0,max(data,na.rm=TRUE)))
lines(data$Minutes, data$Control.1)
lines(data$Minutes, data$Control.2)
lines(data$Minutes, data$Treated)
lines(data$Minutes, data$Treated.1)
lines(data$Minutes, data$Treated.2)
```

We have managed to display the raw data on a plot, but it is unlikely to win any prizes for being aesthetically pleasing! A couple of obvious modifications would be to add a proper x and y label, and to colour the lines appropriately. To change the labels we can specify the `xlab`, `ylab` and `main` arguments. 

```{r}
plot(data$Minutes, data$Control,type="l",ylim=c(0,max(data,na.rm=TRUE)),col="blue",xlab="Minutes",ylab="Count",main="Variation of counts....")
lines(data$Minutes, data$Control.1,col="blue")
lines(data$Minutes, data$Control.2,col="blue")
lines(data$Minutes, data$Treated,col="red")
lines(data$Minutes, data$Treated.1,col="red")
lines(data$Minutes, data$Treated.2,,col="red")
```


```{r}
values <- data[,-1]
Treated <- c(FALSE,FALSE,FALSE,TRUE,TRUE,TRUE)

values[,Treated]
values[!Treated]
```

```{r}
plot(data[,2], type="n",xlab="Minutes",ylab="")
matlines(data[-1],col=c(rep("steelblue",3),rep("salmon",3)),lty=2)
rowMeans(values[Treated],na.rm=TRUE)
points(rowMeans(values[!Treated],na.rm=TRUE),pch=15)
points(rowMeans(values[Treated],na.rm=TRUE),pch=16)
```

##Interpolating Unknowns

Example data for this example is given in the file `Assay.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Assay.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```


```{r}
model <- lm(data$micrograms~data$Optical.Density)
plot(data$Optical.Density, data$micrograms)
abline(model)

```

```{r}

y <- data$micrograms
x <- data$Optical.Density

predict(lm(y~x))

topredict <- which(is.na(y))

new <- data.frame(x = x[which(is.na(y))])
y[which(is.na(y))] <- predict(lm(y~x), new)

cols <- rep("black",length(x))
cols[topredict] <- "red"
plot(x,y,pch=16,col=cols)

```


##Michaelis-Menten data
Example data for this example is given in the file `Michaelis-Menten data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Michaelis-Menten data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##Correlations

Example data for this example is given in the file `Ozone correlations.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Ozone correlations.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

Pairwsie plots of all variables

```{r}
plot(data)
```

```{r}
cor(data)
```



```{r}
cor(data,use="c")
```

```{r}

plot(data$Ozone, data$Wind)
model <- lm(data$Wind ~ data$Ozone)
summary(model)
abline(model,col="red")
```

##Enzyme kinetics - Competitive inhibition


Example data for this example is given in the file `Michaelis-Menten.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Michaelis-Menten.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##Dose Response - x is log(dose)

Example data for this example is given in the file `log-dose vs response.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/log-dose vs response.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we 
print out the first few rows using `head` and print the dimensions
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```


##Dose Response - X is dose

###Data exploration


Example data for this example is given in the file `dose vs.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/dose vs.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

We will now look to visualise the data in R. 

```{r}
plot(data[,2], type="n",axes=FALSE,xlab="log10 Agonist",ylab="")
matlines(data[-1],col=c(rep("steelblue",3),rep("salmon",3)),lty=2)
axis(1,at=1:nrow(data), labels=round(log10(data[,1]),2),)
axis(2)

```

```{r}
values <- data[,-1]
inhibitor <- c(FALSE,FALSE,FALSE,TRUE,TRUE,TRUE)

values[,inhibitor]
values[!inhibitor]

```

```{r}
plot(data[,2], type="n",axes=FALSE,xlab="log10 Agonist",ylab="")
matlines(data[-1],col=c(rep("steelblue",3),rep("salmon",3)),lty=2)
rowMeans(values[inhibitor],na.rm=TRUE)
points(rowMeans(values[!inhibitor],na.rm=TRUE),pch=15)
points(rowMeans(values[inhibitor],na.rm=TRUE),pch=16)
axis(1,at=1:nrow(data), labels=round(log10(data[,1]),2),)
axis(2)

```


##Dose-response - EC50 shift

Example data for this example is given in the file `EC50 shift.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/EC50 shift.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##Binding - Saturation binding to total and non-specific

Example data for this example is given in the file `Total and ns binding data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Total and ns binding data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##Binding - specific binding 

Example data for this example is given in the file `Specific binding data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Specific binding data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```


##RIA or ELISA

Example data for this example is given in the file `RIA data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/RIA data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##Schild data
Example data for this example is given in the file `Schild data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Schild data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```


##Operational model

Example data for this example is given in the file `Operatonal model data.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/Operatonal model data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##XY barplot
Example data for this example is given in the file `XYbarplot.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/XYbarplot.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```

##XY - Frequency

Example data for this example is given in the file `XYFrequency.txt`
```{r echo=FALSE}
myfile <- "PrismExamples/XY/XYFrequency.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We know that the input file is *tab-delimited*, so can use the `read.delim` function. Once the data are read, we print out the first few rows using `head` and print the dimensions.
```{r}
data <- read.delim(myfile,as.is=TRUE)
head(data)
dim(data)
```


# Column data

##Frequency distribution data

Example data for this section is given in the Frequency distribution data.txt

```{r echo=FALSE}
myfile <- "PrismExamples/Column/Frequency distribution data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```


```{r}

data <- read.delim(myfile)
```

We can make a histogram using the `hist` function

```{r}
hist(data$Frequency,col="black",main="Frequency distribution data")
```


```{r}
plot(data$Frequency,pch=16,main="Frequency distribution data")
```

```{r}
plot(data$Frequency,pch=16,main="Frequency distribution data",type="b")
```

##Unpaired t test

The data for this example can be found in the file *Unpaired t test data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Unpaired t test data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

The data consist of `r nrow(data)` rows and `r ncol(data)` columns. The `t.test` function can be to test the null hypothesis that the two groups have the same mean value. The default is to assume that the groups are independent. 

```{r}
t.test(data$Male, data$Female)
```

We can use a boxplot to visualise the data. 


```{r}
boxplot(data)
```

Sometimes we might want to present the observations as points on the graph. We can do this using the `points` function

```{r}
boxplot(data)
points(rep(1, nrow(data)), data[,1],pch=16)
points(rep(2, nrow(data)), data[,2],pch=16)

```

##Paired t test


The data for this example can be found in the file *Paired t test data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Paired t test data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

```{r}
t.test(data$Control, data$Treated,paired=TRUE)
```

```{r}
boxplot(data[,-1])
```

```{r}
boxplot(data[,-1],horizontal = TRUE)
```

##t-test one-sample


The data for this example can be found in the file *One sample t test data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/One sample t test.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##One-way ANOVA
The data for this example can be found in the file *One-way ANOVA data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/One-way ANOVA data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##One-way ANOVA repeated measures

The data for this example can be found in the file *Repeated measures one-way ANOVA data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Repeated measures one-way ANOVA data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```


##ROC Curve

The data for this example can be found in the file *ROC curve data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/ROC curve data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Bland-Altman
The data for this example can be found in the file *Bland-Altman.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Bland-Altman.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Correlation Matrix

The data for this example can be found in the file *Correlation.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Correlation.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Meta-analysis

The data for this example can be found in the file *Metanalysis.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Metanalysis.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Column Statistics

The data for this example can be found in the file *Plasma level of GP-14858.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Column/Plasma level of GP-14858.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

#Grouped

##Two-way ANOVA, not RM

The data for this example can be found in the file *Two-way ANOVA , not RM.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Grouped/Two-way ANOVA , not RM.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Two-Way ANOVA with RM by columns

The data for this example can be found in the file *Two-way ANOVA with RM by columns.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Grouped/Two-way ANOVA with RM by columns.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```



##Two-Way ANOVA with RM by rows

The data for this example can be found in the file *Two-way ANOVA with RM by rows.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Grouped/Two-way ANOVA with RM by rows.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```


##Grouped Bar Graph

The data for this example can be found in the file *First look grouped data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Grouped/First look grouped data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

##Two-way ANOVA ordinary

The data for this example can be found in the file *Two-way ANOVA , not RM.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Grouped/Two-way ANOVA , not RM.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

#Contingency

##Prospective
The data for this example can be found in the file *Prospective (aspirin and MI).txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Contingency//Prospective (aspirin and MI).txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

```{r}
chisq.test(data[,-1])
```


##Retrospective
The data for this example can be found in the file *Retrospective (smoking and cancer).txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Contingency/Retrospective (smoking and cancer).txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

```{r}
 chisq.test(data[,-1])
```



##Sensitivity and Specificity
The data for this example can be found in the file *Sensitivity and specificity (HIV).txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Contingency/Sensitivity and specificity (HIV).txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```


```{r}
fisher.test(data[,-1])
```


##Trend data
The data for this example can be found in the file *Trend data.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Contingency/Trend data.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

```{r}


?prop.trend.test
```



#Survival

##Comparing two groups


The data for this example can be found in the file *Two groups.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Survival//Two groups.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```

Survival analysis is performed by a specialist package `survival` which is part of the standard R distribution. The first task is therefore to load this package, which you only have to do once per R-session.


```{r}
library(survival)
```

The data from Prism require some manipulation before we can apply the survival analysis. Specifically, we need to create *vectors* that contain data about the Time to event, Event and Group. The Time variable

```{r}
Time <- data$Days.elapsed

Group <- rep("Control", nrow(data))
Group[which(is.na(data$Control))] <- "Treatment"
Group

Event <- data$Control
Event[which(is.na(data$Control))] <- data$Treated[which(is.na(data$Control))]
Event
```

The time and data need to be combined into a single vector

```{r}
survData <- Surv(Time, Event)
survData
```


```{r}
plot(survfit(survData~Group))

plot(survfit(survData~Group),col=c("red","blue"),xlab="Days elapsed", 
     main="Survival proportions: Survival of Two groups",
     ylab="Percent survival")

```


```{r}
survdiff(survData~Group)

```


##Comparing three groups


The data for this example can be found in the file *Two groups.txt*. The first step is to locate the file on your disk using the `file.choose()` command and store the result as a variable. 


```{r echo=FALSE}
myfile <- "PrismExamples/Survival//Three groups.txt"
```

```{r eval=FALSE}
myfile <- file.choose()
```

We can now read the file using the `read.delim` function.

```{r}
data <- read.delim(myfile)
data
```


The data from Prism require some manipulation before we can apply the survival analysis. Specifically, we need to create *vectors* that contain data about the Time to event, Event and Group. The Time variable can simply be assigned as the `Days` column. However, to get the other two vectors requires a bit more effort.

```{r}
Time <- data$Days

Group <- rep("Control", nrow(data))
Group[which(!is.na(data$Treatment.A))] <- "TreatmentA"
Group[which(!is.na(data$Treatment.B))] <- "TreatmentB"

Group

Event <- data$Control
Event[which(!is.na(data$Treatment.A))] <- data$Treatment.A[which(!is.na(data$Treatment.A))]
Event[which(!is.na(data$Treatment.B))] <- data$Treatment.B[which(!is.na(data$Treatment.B))]
Event
```



The time and data need to be combined into a single vector

```{r}
survData <- Surv(Time, Event)
survData
```


```{r}
plot(survfit(survData~Group))

plot(survfit(survData~Group),col=c("red","blue","green"),xlab="Days elapsed", 
     main="Survival proportions: Survival of Three groups",
     ylab="Percent survival")

```


```{r}
survdiff(survData~Group)

```
