---
title: "Customising a Plot"
author: "Mark Dunning"
date: "20/11/2014"
output: beamer_presentation
---

#Outline


# Changing how a plot is created

## Specifying extra arguments to plot

- The `plot` function creates a very basic plot
- Many optional arguments can be specified See `?plot`

##Lets re-visit the ozone dataset

The default plots are ugly; No title, Meaningless labels, No colour

```{r fig,width=4,fig.height=4}
data <- read.csv("data/ozone.csv")
plot(data[,1],data[,2])
```


##Use of colours

```{r}
plot(data[,1],col="red")
```

##Plotting characters

```{r}
plot(data[,1], pch=16)
```

##Size of points

```{r}
plot(data[,1], cex=2)
```

```{r}
plot(data[,1], cex=0.2)
```


##Adding a title

```{r}
plot(data[,1], 
     main="Relationship between ozone level and Solar")
```


##Axis labels

```{r}
plot(data[,1], xlab="Ozone level")
```

##Axis labels

```{r}
plot(data[,1], ylab="Solar")
```

##Axis limits

```{r}
plot(data[,1], ylim=c(50,150))
```


##Multiple options at the same time

```{r eval=FALSE,fig.width=4,fig.height=4}
plot(data[,1], pch=16,col="red",
     main="Relationship between ozone level and Solar",
     xlab="Ozone level",
     ylab="Solar")

```

##Multiple options at the same time

```{r echo=FALSE,fig.width=8,fig.height=4}
plot(data[,1], pch=16,col="red",
     main="Relationship between ozone level and Solar",
     xlab="Ozone level",
     ylab="Solar")

```


##Applicable to other types of plot

```{r eval=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
data
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col=c("red","blue","green"))
```

##Applicable to other types of plot

```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col=c("red","blue","green"))
```



# Modifying an existing plot

##Initial plot

```{r}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
```

##The points function

- `points` can be used to set of points to an *existing* plot
- it requires a vector of x and y coordinates
- Note that axis limits of the existing plot are not altered

##Adding points

```{r}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind)
```

##Adding points

`points` can also use the `pch`, `col` arguments. Useful for distinguishing between variables

```{r fig.height=4,fig.width=8}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind,pch=15,col="red")
```


##Adding points

- Each set of points can have a different colour and shape
- Axis labels and title and limits are defined by the plot
- You can add points ad-nauseum. Try not to make the plot cluttered!
- A call to `plot` will start a new graphics window

```{r fig.height=4,fig.width=8}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind,pch=15)
points(data$Ozone, data$Temp,pch=17)
```



##Adding points

- Be careful about the order in which you add points

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16)
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
```

##Adding points

- Can define suitable axis limits in initial plot

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16,ylim=c(0,350))
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
```






## Adding a legend

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16,ylim=c(0,350))
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
legend("topright", legend=c("Solar","Wind","Temp"), col="black", pch=c(16,15,17))
```



# Combining multiple plots

