---
title: "Customising a Plot"
author: "Mark Dunning"
date: "12/12/2014"
output: beamer_presentation
---

# Changing how a plot is created

## Specifying extra arguments to plot

- The `plot` function creates a very basic plot
- Many optional arguments can be specified See `?plot`

##Lets re-visit the ozone dataset

The default plots are ugly; No title, Meaningless labels, No colour

```{r fig,width=4,fig.height=4}
data <- read.csv("data/ozone.csv")
plot(data[,1],data[,2])
```


##Adding a title

```{r}
plot(data[,1], 
     main="Relationship between ozone level and Solar Radiation")
```


##Axis labels

```{r}
plot(data[,1], xlab="Ozone level")
```

##Axis labels

```{r}
plot(data[,1], ylab="Solar Radiation")
```

##Axis limits

```{r}
plot(data[,1], ylim=c(50,150))
```

##Defining a colour

- R can recognise various strings `red`, `orange`,`green`,`blue`,`yellow`....
- Or more exotic ones `r sample(colours(),8)`..... See `colours()`.
- See http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
- Can also use **R**ed **G**reen **Blue** values

##Use of colours

Changing the `col` argument to `plot` changes the colour that the points are plotted in

```{r}
plot(data[,1],col="red")
```


##Plotting characters

- R can use a variety of *p*lotting *ch*aracters
- Each of which has a numeric *code* 
```{r}
plot(data[,1], pch=16)
```

##Plotting characters

```{r echo=FALSE}
par(mar=c(0.1,0.1,0.1,0.1))
i <- 0:24

x <- floor(i /5) + 1
y <- i %%5

plot(1:10, type="n", xlim = c(1,5), ylim=c(-1,5),axes=F,xlab="",ylab="")
points(x,y,pch=i+1)
text(x,y-0.3,i+1)
```


##Plotting characters
- Or you can specify a character

```{r}
plot(data[,1], pch="X")
```


##Size of points

```{r}
plot(data[,1], cex=2)
```

##Size of points

```{r}
plot(data[,1], cex=0.2)
```


##Multiple options at the same time

```{r eval=FALSE,fig.width=4,fig.height=4}
plot(data[,1], pch=16,col="red",
     main="Relationship between ozone level and Solar",
     xlab="Ozone level",
     ylab="Solar")

```

##Multiple options at the same time

```{r echo=FALSE,fig.width=8,fig.height=4}
plot(data[,1], pch=16,col="red",
     main="Relationship between ozone level and Solar",
     xlab="Ozone level",
     ylab="Solar")

```



##Applicable to other types of plot

```{r eval=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
data
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col="red")
```

##Applicable to other types of plot

```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col="red")
```

##What about multiple colours?

- The `col`, `pch` and `cex` arguments are *vectors*
- Previously we used a vector of length one that was *recycled*

```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
```


```{r fig.width=8,fig.height=4}
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col=c("red","blue","green"))
```


##Don't get carried away

- Each point can have a unique colour, plotting character, size.

```{r echo=FALSE}
data <- read.csv("data/ozone.csv")
```


```{r echo=FALSE}
plot(data[,1], pch=1:20, col=sample(colours(), nrow(data)))
```

##Can modify specific points

- Suppose we know that observations 117, 62, 99, 121 and 30 were the highest ozone level
- We may wish to plot them a different colour
- *a* Solution: Create a vector of colours the required length and modify the appropriate entries

```{r fig.width=8,fig.height=4}

mycols <- rep("black", 153)
mycols[c(117,62,99,121,30)] <- "red"

plot(data[,1], pch=16, col=mycols)
```


##Using a palette
- The `RColorBrewer` package has various ready-made colour schemes
```{r}
library(RColorBrewer)
display.brewer.all()
```

##Creating a palette

- `brewer.pal` function creates a vector of the specified length comprising colours from the named palette

```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
```


```{r fig.height=4,fig.width=8}

mypal <- brewer.pal(3, "Set1")
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col=mypal)
```



# Modifying an existing plot

##Initial plot

```{r}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
```

##The points function

- `points` can be used to set of points to an *existing* plot
- it requires a vector of x and y coordinates
- Note that axis limits of the existing plot are not altered

##Adding points

```{r}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind)
```

##Adding points

`points` can also use the `pch`, `col` arguments. Useful for distinguishing between variables

```{r fig.height=4,fig.width=8}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind,pch=15,col="red")
```


##Adding points

- Each set of points can have a different colour and shape
- Axis labels and title and limits are defined by the plot
- You can add points ad-nauseum. Try not to make the plot cluttered!
- A call to `plot` will start a new graphics window

```{r fig.height=4,fig.width=8}
data <- read.csv("data/ozone.csv")
plot(data$Ozone, data$Solar.R,pch=16)
points(data$Ozone, data$Wind,pch=15)
points(data$Ozone, data$Temp,pch=17)
```



##Adding points

- Be careful about the order in which you add points

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16)
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
```

##Adding points

- Can define suitable axis limits in initial plot

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16,ylim=c(0,350))
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
```




## Adding a legend

```{r fig.height=4,fig.width=8}
plot(data$Ozone, data$Wind,pch=16,ylim=c(0,350))
points(data$Ozone, data$Solar.R,pch=15)
points(data$Ozone, data$Temp,pch=17)
legend("topright", legend=c("Solar","Wind","Temp"), 
       col="black", pch=c(16,15,17))
```

##Adding text

```{r fig.height=4,fig.width=8}
mycols <- rep("black", 153)
mycols[c(117,62,99,121,30)] <- "red"

plot(data[,1], pch=16, col=mycols)
text(c(117,62,99,121,30), data[c(117,62,99,121,30),1], 
     labels=LETTERS[1:5])
```

##Adding lines

```{r fig.height=4,fig.width=8}
mycols <- rep("black", 153)
mycols[c(117,62,99,121,30)] <- "red"

plot(data[,1], pch=16, col=mycols)
abline(h = 115)

```



##Adding lines

```{r}

plot(data[,1], pch=16, col=mycols)
grid(col="steelblue")
```

##Adding lines

- `abline` can take a gradient and intercept argument
- Can be used to draw a *line of best fit* in conjunction with a linear model

```{r}
plot(1:10, jitter(1:10))
abline(0,1)
```


##Adding lines

Lines can also be added to other plots
```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.delim("data/plasma.txt")
```

```{r fig.height=4,fig.width=8}
boxplot(data, main="Cell counts",xlab="Cell type",
        ylab="Count",col="red")
abline(h=c(4,5,6),col="steelblue")
```

##Adding lines

Lines can also be added to other plots

```{r fig.height=4,fig.width=8}
barplot(colMeans(data))
abline(h=c(4,5,6),col="steelblue")
```

##See also

- rect `example(rect)`
- polygon `example(polygon)`
- segments `example(segments)`


# Plot layout options

##The `par` function

- Using the `par` function prior to creating a plot allows several plot defaults to be set
- ?par for details

```{r}

```


##Multiple figures

- We can have *m*ultiple *f*igures per *row* using `mfrow`
- e.g. one row and three columns
- each new call to plot is added in a new panel
- see also `mfcol`

```{r echo=FALSE,fig.width=8,fig.height=4}
data <- read.csv("data/ozone.csv")
```

```{r fig.height=4,fig.width=8}
par(mfrow=c(1,3))
plot(data[,1],data[,2])
plot(data[,1],data[,3])
plot(data[,1],data[,4])

```

##Margin size

- the `mar` vector specifies that amount of space around each edge of the plot
-  `c(bottom, left, top, right)`


```{r echo=FALSE,results='hide'}
dev.off()
```


```{r echo=FALSE,fig.width=8,fig.height=4}

par(mar = c(5,1,5,1))
plot(data[,1],data[,2])


```

#Exporting a plot

##As a png
- `png` function prior to code to create plot
- file is created in your working directory (doesn't need to exist)
- `dev.off()` afterwards
- can also make jpeg in similar fashion
```{r}
png("mycoolplot.png")
plot(data[,1],data[,2])
dev.off()
```


##As a pdf

- As before, except use `pdf`

```{r}
pdf("mycoolplot.pdf")
plot(data[,1],data[,2])
dev.off()
```

##As a pdf

- However, a pdf can have multiple pages
- Can editted by program such as Photoshop
- Can specify dimensions, dpi etc
```{r}
pdf("mycoolmultipageplot.pdf")
plot(data[,1],data[,2])
plot(data[,1],data[,3])
dev.off()
```

#Break for practical