---
title: "Plotting Basics"
author: "Mark Dunning"
date: "12/12/2014"
output: beamer_presentation
---


#The scatter plot

## Introducing the dataset
These data were obtained from a [Guardian blog of June 2011](http://www.theguardian.com/news/datablog/2011/jun/08/life-expectancy-uk-data-health).

```{r}
birth <- read.delim("data/UKBirthRate.tsv")

```

## Exploring the data

You should *always check* that the data have been imported correctly.
```{r}
head(birth)
dim(birth)
```

## Exploring the data

```{r}
summary(birth)
str(birth)
```

##Data representation

The data are stored in a data frame. These are subset using square brackets `[]`

```{r}
birth[1:10,1]
```

##Data representation

We can get particular columns and rows by *omitting* the row or column index. The result is a `vector`

```{r}
birth[1,]
birth[,1]
```

##Data representation

The data frame is *not altered*

```{r}
dim(birth)
birth[1,]
dim(birth)
```

##Assignment to a variable

Again, the result is a `vector`
```{r}
mBirth <- birth[,2]
mBirth
```

##Assignment to a variable

Note that we can access columns by *name* with the `$` operator

```{r}
mBirth <- birth$Male.babies
mBirth
```

##Scatter plots 

Suppose we want to look at the change in male birth rate over the years. `plot` is the general-purpose plotting function in R

```{r}
plot(mBirth)
```

##Data visualisation

```{r}
years <- birth$Age
plot(years, mBirth)
```

##Data visualisation

Can plot one vector against another
```{r}
fBirth <- birth$Female.babies

plot(mBirth, fBirth)
```

#Visualising Distributions

##The dataset

Describes weather conditions on different days over a period of months

```{r echo=FALSE}

data <- read.delim("FromFran/hAREG.txt")
head(data)
```

##The boxplot

Can use the `~` sign to signify the grouping of the data

```{r}
boxplot(data)
```

##Plotting individual points

```{r}
stripchart(data,vertical=TRUE)

```


##Plotting individual points

```{r}
library(beeswarm)
beeswarm(data,vertical=TRUE)

```

##Overlaying points

```{r}
boxplot(data)
beeswarm(data,vertical=TRUE,add=TRUE)
```

##Summarising the data

```{r}
colMeans(data,na.rm=TRUE)
mns <- colMeans(data,na.rm=TRUE)
sds <- apply(data, 2, sd,na.rm=TRUE)
```


##Bar plots

```{r}
bp <- barplot(colMeans(data,na.rm=TRUE),ylim=c(0,400))
arrows(bp, mns + sds,bp, mns,code=3,angle=90)
```


#Count data

#Curves