---
title: "Plotting Basics"
author: "Mark Dunning"
date: "12/12/2014"
output: beamer_presentation
---


#The scatter plot

## Introducing the dataset
These data were obtained from a [Guardian blog of June 2011](http://www.theguardian.com/news/datablog/2011/jun/08/life-expectancy-uk-data-health).

```{r}
birth <- read.delim("data/UKBirthRate.tsv")

```

## Exploring the data

You should *always check* that the data have been imported correctly.
```{r}
head(birth)
dim(birth)
```

## Exploring the data

```{r}
summary(birth)
str(birth)
```

##Data representation

The data are stored in a data frame. These are subset using square brackets `[]`

```{r}
birth[1:10,1]
```

##Data representation

We can get particular columns and rows by *omitting* the row or column index. The result is a `vector`

```{r}
birth[1,]
birth[,1]
```

##Data representation

The data frame is *not altered*

```{r}
dim(birth)
birth[1,]
dim(birth)
```

##Assignment to a variable

Again, the result is a `vector`
```{r}
mBirth <- birth[,2]
mBirth
```

##Assignment to a variable

Note that we can access columns by *name* with the `$` operator

```{r}
mBirth <- birth$Male.babies
mBirth
```

##Thinking about the data
What variables do we have?

- Year (Continuous)
- Birth Rate (Continuous)
- Gender (Discrete)

What are we interested in?

- Trend
- Relationship

Any repeated observations?

- No

##Scatter plots 

Suppose we want to look at the change in male birth rate (continuous)

- `plot` is the general-purpose plotting function in R

```{r}
plot(mBirth)
```

##Data visualisation
male birth rate (continuous) versus years (continuous)

```{r}
years <- birth$Age
plot(years, mBirth)
```

##Data visualisation

Can plot one vector against another
```{r}
fBirth <- birth$Female.babies

plot(mBirth, fBirth)
```

##Adding lines and points

#Visualising Distributions

##The dataset



##The boxplot


##Plotting individual points


##Plotting individual points



##Overlaying points

##Summarising the data



##Bar plots



#Count data

#Curves