---
title: "Plotting Basics"
author: "Mark Dunning"
date: "12/12/2014"
output: beamer_presentation
---


#The scatter plot

## Introducing the dataset
These data were obtained from the [supplementary data](http://faculty.washington.edu/heagerty/Books/Biostatistics/index-chapter.html) to *Biostatistics: A Methodology for the Health Sciences*


```{r echo=FALSE}
if(!file.exists("data/ozone.csv")) download.file(url = "http://faculty.washington.edu/heagerty/Books/Biostatistics/DATA/ozone.csv", destfile = "data/ozone.csv")
```

```{r}
data <- read.csv("data/ozone.csv")
```

## Exploring the data

You should *always check* that the data have been imported correctly.
```{r}
head(data)
dim(data)
```

## Exploring the data

```{r}
summary(data)
str(data)
```

##Data representation

The data are stored in a data frame. These are subset using square brackets `[]`

```{r}
data[1:10,1]
```

##Data representation

We can get particular columns and rows by *omitting* the row or column index. The result is a `vector`

```{r}
data[1,]
data[,1]
```

##Data representation

The data frame is *not altered*

```{r}
dim(data)
data[1,]
dim(data)
```

##Assignment to a variable

Again, the result is a `vector`
```{r}
ozone <- data$Ozone
```


##Thinking about the data
What variables do we have?

- Ozone, Wind, Temp (Continuous)
- Month, Day (Discrete)

What are we interested in?

- Trend
- Relationship

Any repeated observations?

- No

##Scatter plots 

Suppose we want to look at the change in Ozone level (continuous)

- `plot` is the general-purpose plotting function in R
- Given a *vector* it will plot the values in the vector on the **y** axis, and index on the **x** axis

```{r fig.width=8,fig.height=4}
plot(data$Ozone)
```

##Scatter plots 

- We have `r nrow(data)` points on the plot
- Axis labels, points, title and colours can be altered

```{r fig.width=8,fig.height=4}
plot(data$Ozone)
```


##Data visualisation

- Can plot one vector against another
- First *argument* is plotted on the x axis, second *argument* on the y axis

```{r fig.width=8,fig.height=4}
plot(data$Ozone,data$Wind)
```

##Adding lines

```{r fig.width=8,fig.height=4}
plot(data$Ozone)
abline(v= c(50,100,150))
abline(h=100)
```

##Adding points

- Suppose we are interested in days with Ozone level over 100
- We can form a *logical* vector and get the `TRUE` values using the `which` function

```{r}
highOzone <- which(data$Ozone > 100)
data[highOzone,]
```

##Adding points

- The `points` funtion is used to add points to an existing plot
- We need to give it a set of *x* and *y* coordinates
- The *x* values are the indices we've jsut computed. 
- *y* values are obtained by subsetting the *Ozone* variable

```{r}
newX <- highOzone
newY <- data$Ozone[newX]
```

##Adding points

- We use a different colour and character for the points. More on this later

```{r fig.width=8,fig.height=4}
highOzone <- which(data$Ozone > 100)
plot(data$Ozone)
abline(h=100)
points(newX, newY,col="red",pch=16)
```



##Multiple comparisons

- The `plot` function will try and *guess* what plot to make based on the data
- If given a *data frame* it make pairwise comparisons of all columns

```{r}
plot(data[,1:4])
```

##Exercises

- Read the tab-delimited version of the weather data and check the dimensions
- Plot the *Temperature* variable
- Which days have a temperature > 90
- Highlight these days on the scatter plot
- Plot the Temperature against the Ozone level

##Answers

```{r}
data2 <- read.delim("data/ozone.txt")
```

```{r fig.width=8,fig.height=4}
plot(data$Temp)
```

##Answers

```{r fig.width=8,fig.height=4}
highTemp <- which(data$Temp > 90)
plot(data$Temp)
newX <- highTemp
newY <- data$Temp[highTemp]
abline(h=90)
points(newX,newY,col="red",pch=16)
```

##Comments
- Don't need to create `newX` and `newY` variables
```{r fig.width=8,fig.height=4}
plot(data$Temp)
abline(h=90)
points(highTemp, data$Temp[highTemp],col="red",pch=16)
```

##Comments

- Better not to *hard-code* the limit

```{r fig.width=8,fig.height=4}
tempLim <- 95
highTemp <- which(data$Temp > tempLim)
plot(data$Temp)
abline(h=tempLim)
points(highTemp, data$Temp[highTemp],col="red",pch=16)

```

##Answers

```{r}
plot(data$Temp,data$Ozone)
```

##Comments

- The variables seem to be related. We can quantify the correlation and test for significance
```{r}
cor(data$Temp,data$Ozone,use="c")
cor.test(data$Temp,data$Ozone,use="c")
```

#Visualising Distributions

##The dataset

```{r}
data <- read.delim("data/plasma.txt")
data
```



##The boxplot

```{r fig.width=8,fig.height=4}
boxplot(data)
```


##Plotting individual points

- `vertical = TRUE` ensures the plot is in the same orientation as the boxplot

```{r fig.width=8,fig.height=4}
stripchart(data,vertical=TRUE)
```

##Plotting individual points

- We can *stack* or *jitter* points if required

```{r fig.width=8,fig.height=4}
stripchart(data,vertical=TRUE,method="jitter")
```






##Plotting individual points
```{r}
library(beeswarm)
beeswarm(data,vertical=TRUE)
```



##Overlaying points

- `add=TRUE` argument overlays the stripchart on an existing plot

```{r fig.width=8,fig.height=4}
boxplot(data)
stripchart(data,vertical=TRUE,add=TRUE)

```


##Summarising the data

```{r}
summary(data)
```


##Bar plots

```{r}
barplot(colMeans(data))
```


##Adding error bars

```{r echo=FALSE}
dpPlot <- function(data){
  
  mns <- colMeans(data,na.rm=TRUE)
  sds <- apply(data, 2, sd, na.rm=TRUE)
  
  err <- mns + sds
  xs <- barplot(mns,ylim=c(0, max(err)),axes=F)
  
  arrows(xs, mns, xs, err,code=3,angle=90)
  axis(2,at=pretty(0:floor(max(err))))
}
```

```{r}
dpPlot(data)
```


##What to do with long data

- Recall our weather data
- We do not have separate columns for each month

```{r}
data <- read.csv("data/ozone.csv")
head(data)
```

##Boxplot of long data

- Month is a variable in the data frame
- We use *formula* syntax with the `~` symbol

```{r}
boxplot(data$Ozone~data$Month)
```

##Boxplot of long data

```{r}
boxplot(data$Temp~data$Month)
```

##stripchart of long data

```{r}
stripchart(data$Ozone~data$Month,vertical=TRUE)
```

##Boxplot of long data

```{r}
boxplot(data$Ozone~data$Month)
stripchart(data$Ozone~data$Month,vertical=TRUE,add=TRUE)
```

##Boxplot of long data

- This is equivalent and a bit more concise
```{r}
boxplot(Ozone~Month,data)
stripchart(Ozone~Month,data,vertical=TRUE,add=TRUE)
```

##Melting a dataset

We can convert a dataset from *wide* to *long* using the `reshape` package

```{r}
library(reshape2)
data <- read.delim("data/plasma.txt")
plasma <- melt(data)
dim(plasma)
head(plasma)
```

##Boxplot of the new data

- Compare to before...

```{r}
boxplot(value ~ variable,plasma)
```



#Count data

##Making a barplot

```{r fig.width=8,fig.height=4}
clinical <- read.delim("data/NKI295.pdata.txt")
table(clinical$ER)
barplot(table(clinical$ER))
```

##Making a barplot
```{r fig.width=4,fig.height=4}
pie(table(clinical$ER))
```

##Stacking 

```{r fig.width=8,fig.height=4}
table(clinical$ER,clinical$grade)
barplot(table(clinical$grade,clinical$ER))
```

##Stacking 

```{r fig.width=8,fig.height=4}
table(clinical$ER,clinical$grade)
barplot(table(clinical$grade,clinical$ER),beside=TRUE)
```

#Curves

##Survival curves

To perform a survival analysis we need the following pieces of information

- Time to Event
- Event (e.g. dead or alive)
- Group

##Example data

```{r}
clinical <- read.delim("data/NKI295.pdata.txt")

Event <- clinical$event_death
Time <- clinical$survival.death.
Group <- clinical$ER

```

##The survival package

```{r}
library(survival)
survData <- Surv(Time, Event)
survData[1:10]
```


##Making the Survival curve

```{r}
plot(survfit(survData ~ Group))
```

##Testing

```{r}
survdiff(survData~Group)
```

##Data from Prism

- Survival data in Prism have an annoying format. Time is easily obtained from the table

```{r}
sData <- read.delim("data/Two groups.txt")
head(sData)
Time <- sData$Days.elapsed
```

##Data from Prism

- However, a bit of fiddling required to get a Event and Group vector

```{r}
Group <- rep("Control", nrow(sData))
Group[which(is.na(sData$Control))] <- "Treatment"
Group
```

##Data from Prism

```{r}
Event <- sData$Control
Event[which(is.na(sData$Control))] <- sData$Treated[which(is.na(sData$Control))]
Event
```

##Making the survival curve

```{r}
survData <- Surv(Time, Event)
plot(survfit(survData~Group))
```

